% $Id$
% Copyright (c) 2000  The PARI Group
%
% This file is part of the PARI/GP documentation
%
% Permission is granted to copy, distribute and/or modify this document
% under the terms of the GNU General Public License
\appendix{Summary of Available Constants}

In this appendix we give the list of predefined constants available in the
PARI library. All of them are in the \idx{heap} and \emph{not} on the PARI
\idx{stack}. We start by recalling the \idx{universal object}s introduced in
\secref{se:intro4}:
%
\bprog
t_INT: gen_0, gen_1, gen_m1, gen_2, gen_m2
t_FRAC: ghalf
@eprog
\noindent Only polynomials in the variables \kbd{0} and \kbd{MAXVARN} are
defined initially. Use \kbd{fetch\_var()} (see \secref{se:fetch_var}) to
create new ones.

\noindent The other objects are not initialized by default:

\tet{bern}(i). This is the $2i$-th Bernoulli number ($B_0=1$, $B_2=1/6$,
$B_4=-1/30$, etc\dots). To initialize them, use \tet{mpbern}.

In addition, some single or double-precision real numbers (like \kbd{PI}) are
predefined, and their list is in the file \kbd{paricom.h}.

Finally, one has access to a table of (differences of) primes through the
pointer \tet{diffptr}. This is used as follows: when

\fun{void}{pari_init}{size\_t size, ulong maxprime}

\noindent is called, this table is initialized with the successive
differences of primes up to (just a little beyond) \kbd{maxprime}
(see \secref{se:intro4}). The prime table will occupy roughly
$\kbd{maxprime}/\log(\kbd{maxprime})$ bytes in memory, so be sensible when
choosing \kbd{maxprime} (it is $500000$ by default under \kbd{gp}). In any case,
the implementation requires that $\tet{maxprime} < 2^{\B} - 2048$, whatever
memory is available.

\noindent Some convenience functions:

\fun{ulong}{maxprime}{} the largest prime computable using our prime table.

\fun{void}{maxprime_check}{ulong B} raise an error if \kbd{maxprime()} is $< B$.

After the following initializations (the names $p$ and \var{ptr} are arbitrary of
course)
\bprog
byteptr ptr = diffptr;
ulong p = 0;
@eprog
\noindent calling the macro \tet{NEXT_PRIME_VIADIFF_CHECK}$(p, \var{ptr})$
repeatedly will assign the successive prime numbers to $p$. Overrunning the
prime table boundary will raise the error \tet{primer1}, which will just
print the error message:

\kbd{*** not enough precomputed primes}

\noindent and then abort the computations. The alternative macro
\tet{NEXT_PRIME_VIADIFF} operates in the same way, but will omit that check, and
is slightly faster. It should be used in the following way:
%
\bprog
byteptr ptr = diffptr;
ulong p = 0;

if (maxprime() < goal) pari_err(primer1); /*@Ccom not enough primes */
while (p <= goal) /*@Ccom run through all primes up to \kbd{goal} */
{
  NEXT_PRIME_VIADIFF(p, ptr);
  ...
}
@eprog\noindent
Here, we use the general error handling function \kbd{pari\_err} (see
\secref{se:err}), with the codeword \kbd{primer1}, raising the ``not enough
primes'' error. This could be rewritten as
\bprog
maxprime_check(goal);
while (p <= goal) /*@Ccom run through all primes up to \kbd{goal} */
{
  NEXT_PRIME_VIADIFF(p, ptr);
  ...
}
@eprog

\fun{bytepr}{initprimes}{ulong maxprime} computes a prime table (of all prime
differences for $p < maxp$) on the fly. You may assign it to \kbd{diffptr} or
to a similar variable of your own. Beware that before changing \kbd{diffptr},
you should really free the (\kbd{malloc}ed) precomputed table first; and then
all pointers into the old table will become invalid.

PARI currently guarantees that the first 6547 primes, up to and including
65557, are present in the table, even if you set \kbd{maxprime} to zero.
in the \kbd{pari\_init} call.

\fun{ulong}{init_primepointer}{ulong a, ulong p, byteptr *ptr}
assume \kbd{*ptr} is inside of a \kbd{diffptr} containing the succesive
differences between primes, and $p$ is the current prime (up to \kbd{*ptr}
excluded). Returns return the smallest prime $\geq a$, and update \kbd{ptr}.

\vfill\eject
